Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model

### 2022-01-10 manuscripts/owp_fim4_2021/sections/abstract_plainTextSummary.tex
edits after odgen/salas comments

\begin{abstract}
The National Water Model Version 2.1 (NWM V2.1) seeks to expand upon existing systems to provide hourly forecast discharges at four time horizons (1, 3, 10, and 30 days) at over 2.7 million locations throughout the continental United States, Hawaii and Puerto Rico.
Additional modeling capabilities are required to convert these 1-D discharge forecasts into 2-D fluvial inundation extents used for forecasting purposes to protect life and property.
Height Above Nearest Drainage (HAND) coupled with the use of synthetic rating curves can convert discharges into stages and stages to inundation extents in a rapid manner across large spatial domains and at high resolutions.
The Office of Water Prediction Flood Inundation Mapping (OWP FIM) model `Cahaba' is an enhancement of the HAND method developed to improve upon some of HAND's limitations.
Among the implemented enhancements are efficient computational performance, support for Hawaii and Puerto Rico NWM domains, enforcement of levee elevation data, support for the latest digital elevation model and stream network data sources, advanced stream burning techniques, thalweg conditioning, and simplified Manning's n variation.
Most importantly, we detail an issue previously mentioned with HAND as its FIM skill is dependent on a drainage threshold parameter used to delineate stream lines.
Since our modeling assumptions are different, we illustrate how reducing Horton-Strahler stream order of the stream networks used within each HAND processing unit down to unary can improve FIM skill.
This is accomplished by deriving level paths for the National Water Model stream network and utilizing those definitions for deriving HAND datasets independently from other level paths.
`Cahaba' was evaluated at 49 Base Level Engineering (BLE) sites furnished by the Federal Emergency Management Agency (FEMA).
The modeled discharges from the BLE sections were used for the 100 and 500 year flood events as input to OWP FIM instead of NWM discharges to isolate out factors related to hydro-meteorology.
The resulting maps are compared to the modeled extents from the BLE and binary contingency statistics are computed.
Across the 49 sites, Critical Success Index (CSI) was found to increase across FIM versions driven by improvements to the Probability of Detection (POD) and slight reduction in False Alarm Ratio (FAR).
Each successive version of HAND here utilizes greater stream order reduction illustrating a key limitation of HAND and enhanced agreement when utilizing stream networks of lower stream orders as a datum for HAND.
Additionally we detail the effects of two different Manning's n values and how order reduction drives rating curve bias downward due to the nature of reach averaged synthetic rating curves.
These capabilities were implemented in an open-source package available on GitHub (\url{https://github.com/NOAA-OWP/cahaba}) for continued use and collaboration with the research community.
%
\end{abstract}
%
\section*{Plain Language Summary}
Flooding is one of the most impactful natural disasters on life and property.
The United States National Water Model (NWM) seeks to provide flood forecasts for the entire country so that adequate warnings can be raised to the public to enable safe evacuations and protective measures.
In order to convert forecasted flow rates from the NWM to flood inundation maps (FIM), a model is used that converts elevation data from height above mean sea-level to height above the nearest river bottom.
This model known as Height Above Nearest Drainage (HAND) suffers from issues in mapping performance where rivers converge.
We developed a technique that mitigates these errors by removing consideration for neighboring tributaries in the relative elevation computation process.
We compared these HAND derived FIMs to maps from more realistic models and found and improvement in mapping performance.
%

>>>>

\begin{abstract}
The National Water Model Version 2.1 (NWM V2.1) seeks to expand upon existing systems to provide up to hourly forecast discharges for four time horizons (1, 3, 10, and 30 days) at over 2.7 million locations throughout the continental United States, Hawaii and Puerto Rico.
Additional modeling capabilities are required to convert these 1-D discharge forecasts into 2-D fluvial inundation extents used for forecasting purposes to aid in the protection of life and property.
Height Above Nearest Drainage (HAND) coupled with the use of synthetic rating curves can convert discharges into stages and stages to inundation extents in a rapid manner across large spatial domains and at high resolutions.
The Office of Water Prediction Flood Inundation Mapping (OWP FIM) model `Cahaba' is an enhancement of the HAND method developed to improve upon some of HAND's limitations.
Among the implemented enhancements are efficient computational performance, support for Hawaii and Puerto Rico NWM domains, enforcement of levee elevation data, support for the latest digital elevation model and stream network data sources, advanced stream burning techniques, thalweg conditioning, and simplified Manning's n variation.
Most importantly, we detail an issue previously mentioned with HAND as its FIM skill is dependent on a drainage threshold parameter used to delineate stream lines.
Since our modeling assumptions are different, we illustrate how reducing Horton-Strahler stream order of the stream networks used within each HAND processing unit down to unary can improve FIM skill.
This is accomplished by deriving level paths for the National Water Model stream network and utilizing those definitions for deriving HAND datasets independently from other level paths.
`Cahaba' was evaluated at 49 Base Level Engineering (BLE) sites furnished by the Interagency Flood Risk Management (InFIRM).
The modeled discharges for 100 and 500 year events from the BLE cross-sections were used for input to OWP FIM instead of NWM discharges to isolate out factors related to hydro-meteorology.
The resulting maps are compared to the modeled extents from the BLE and binary contingency statistics are computed.
Across the 49 sites, Critical Success Index (CSI) was found to increase across FIM versions driven by improvements to the Probability of Detection (POD) and slight reduction in False Alarm Ratio (FAR).
Each successive version of HAND here utilizes greater stream order reduction illustrating a key limitation of HAND and enhanced agreement when utilizing stream networks of lower stream orders as a datum for HAND.
Additionally we detail the effects of two different Manning's n values and how order reduction drives rating curve bias downward due to the nature of reach averaged synthetic rating curves.
These capabilities were implemented in an open-source package available on GitHub (\url{https://github.com/NOAA-OWP/cahaba}) for continued use and collaboration with the research community.
%
\end{abstract}
%
\section*{Plain Language Summary}
Flooding is one of the most impactful natural disasters on life and property.
The United States National Water Model (NWM) provides flood forecasts for the entire country so that adequate warnings can be raised to the public to enable safe evacuations and protective measures.
In order to convert forecasted flow rates from the NWM to flood inundation maps (FIM), a model is used that converts elevation data from height above mean sea-level to height above the nearest river bottom.
This model known as Height Above Nearest Drainage (HAND) suffers from issues in mapping performance where rivers converge.
We developed a technique that mitigates these errors by removing consideration for neighboring tributaries in the relative elevation computation process.
We compared these HAND derived FIMs to maps from more realistic models and found and improvement in mapping performance.
%

### 2022-01-10 manuscripts/owp_fim4_2021/sections/conclusions.tex
edits after odgen/salas comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Overall, we have produced a product that is able to produce FIMs for the entire US using stream flows whether retrospective or forecast.
Using the HAND method, we can produce this product at very large domains and high resolutions in a relatively short amount of time which enables creating forecast maps at high temporal frequencies.
This yields a significant amount of intelligence that was previously unavailable to forecasters.
All of this is presented in an open-source framework that utilizes some of the latest enhancements to HAND as well as some of the latest data sources available. 
HAND is a terrain index that normalizes height above mean sea level to height above the nearest, relevant drainage line thus detrending elevations.
It is dependent on a series of drainage enforcing operations and suffers from limitations at stream junctions that partially result from independent catchments and multi-order stream networks.
This issue is addressed here by reducing the scale at which HAND is derived from a HUC level to one of a levelpath that reduces the stream order down to a unary level.
We illustrated how this technique improves FIM skill especially at junctions of higher order and higher flow magnitude rivers.
We additionally illustrated the sensitivity to rating curve bias and implications of stream order reduction.
Lastly, continental scale flood forecasting still has much progress left to make so we proposed a path forward for improving FIM via the enhancement of terrain data, roughness estimates, and physics constraints. 
%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Floods present a significant, underserved, and expanding risk to life, property, and resources.
Previous flood forecasting systems lacked the coverage to adequately inform society of these risks.
The National Water Model (NWM) developed by the National Oceanic and Atmospheric Administration's Office of Water Prediction, along with partners, provides increased spatial coverage and resolution as well as additional forecast time horizons on mostly hourly intervals.
Additional processing is required to convert streamflows from the NWM to river stages and finally to flood inundation maps (FIM).
Height Above Nearest Drainage (HAND) is a means of detrending digital elevations maps (DEM) by normalizing elevation to the nearest relevant drainage point.
HAND coupled with the use of reach averaged synthetic rating curves (SRC) provide such a means of creating continental scale FIM capabilities at high resolutions (1/3 arc-second, 10 m) and high temporal frequencies (up to 1 hr).
Scalable, open-source software was developed to produce HAND and associated datasets (catchments, SRCs, and cross-walking data) for the NWM forecasting area including Hawaii and Puerto Rico (https://github.com/NOAA-OWP/cahaba).
HAND is produced using the latest hydro-conditioning techniques to enforce monotonically decreasing elevations including stream burning, levee enforcement, pit-filling, stream channel excavation, thalweg breaching, headwater seeding, stream reach resampling, and more. 
Finally, we use this implementation to investigate the skill of the FIMs by varying the scale of the processing units used to derive HAND.
We illustrate that reducing the Horton-Strahler stream order of a HAND processing unit down to one enhances skill by significantly reducing false negatives at junctions of major streams.
This also affects the parameters used to compute stage-discharge relationships biasing discharge higher at given stages which reduced the rate of false positives.
FIM skill was evaluated over large spatial scales by comparsion to HEC-RAS 1D models.
Further investment in the SRC's is warranted by accounting for bathymetric errors inherited by the DEM and better accounting for localized friction values at varying flow magnitudes.
Utilizing the highest resolution Lidar and bathymetric data should also improve the vertical accuracy of HAND and better account for fine grain features that greatly affect inundation extents.
Due to inherent limitations with HAND, scalable physics-based methods will need to be worked on to provide better representation of flood extent dynamics in steady and unsteady conditions. 
%

### 2022-01-10 manuscripts/owp_fim4_2021/sections/discussion.tex
edits after odgen/salas comments

In analyzing the data, we found a slight reduction in FAR was detected and more digging pointed to a bias in rating curves introduced by stream order reduction.
Figure \ref{fig:rating_curve_comparison} illustrates the general effect that stream order reduction has on synthetic rating curves.
Sub-figure \ref{fig:rating_curve_comparison}a shows how the average rating curves for all reaches for stage values 0 to 25 meters at one-third meter intervals tend to bias down (and to the right) with ever increasing stream order reduction (FR to MS to GMS). 
This bias is more pronounced for GMS since that implements stream order reduction down to the unit level for the entire FR network while MS only does so for 4-5\% of the network.
Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged synthetic rating curve relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's n values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V[y] or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B[y] or B) is analogous to reach-averaged hydraulic radius.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively. 
Figures \ref{fig:rating_curve_comparison} b and c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but since both are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ is plotted against stage, y, to show how these two parameters collectively pull the rating curve Q up and biases the rating curve down.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.
Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments that encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily including the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this fact to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

@@

Our qualitative analysis suggests that the synthetic rating curves offer a significant opportunity for improvement in HAND based FIM for future development.
The bathymetry of the NED 10m DEM is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's n which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3m and 1m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.

>>>>

In analyzing the data, we found a slight reduction in FAR was detected and more digging pointed to a bias in rating curves introduced by stream order reduction.
Figure \ref{fig:rating_curve_comparison} illustrates the general effect that stream order reduction has on synthetic rating curves.
Sub-figure \ref{fig:rating_curve_comparison}a shows how the average rating curves for all reaches for stage values 0 to 25 meters at one-third meter intervals tend to bias down (and to the right) with ever increasing stream order reduction (FR to MS to GMS). 
This bias is more pronounced for GMS since that implements stream order reduction down to the unit level for the entire FR network while MS only does so for 4-5\% of the network.
Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged synthetic rating curve relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's n values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V(y) or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B(y) or B) is analogous to reach-averaged hydraulic radius.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively (see Equation \ref{eq:reach_averaged_mannings_equation}). 
Figures \ref{fig:rating_curve_comparison} b and c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ and plot this ratio against stage, y, to show how these two parameters collectively pull the rating curve Q up and biases the rating curve down.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.
Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments that encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily including the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this fact to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

@@

Our qualitative analysis suggests that the synthetic rating curves offer a significant opportunity for improvement in HAND based FIM for future development.
The bathymetry of the 10 m DEM from 3DEP is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's n which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3 m and 1 m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.

### 2022-01-10 manuscripts/owp_fim4_2021/sections/introduction.tex
edits after odgen/salas comments

One of these operational systems is the Advanced Hydrologic Prediction System (AHPS) maintained by National Oceanic Atmospheric Administration (NOAA) National Weather Service (NWS) with thousands of forecasting points across the US at typically short forecast horizons of 24 or 72 hours \cite{mcenery2005noaa}.
AHPS provides forecasting services in the form of ensemble stream flows at more than 3,000 locations and flood inundation maps (FIM) at more than 150 of those points shown in Figure \ref{fig:forecast_points}.

@@

Despite the AHPS advances in operational flood forecasting, it lacks sufficient spatial coverage and long-range forecast horizons.

@@

\caption{Forecast points with and without FIM in United States' Advanced Hydrologic Prediction System. Also show are the National Water Model stream networks at the full resolution (FR) and Replace and Route (RnR) Mainstems (MS) resolution.}

@@

To broaden the forecasting domain, the Office of Water Prediction (OWP) at the National Water Center (NWC) in Tuscaloosa, Alabama commissioned the development of the National Water Model (NWM) which is a version of the Weather Research and Forecast Hydrologic Model (WRF-Hydro) \cite{gochis2018wrf,cosgrove2019evolution}. 

@@

The NWM enhances but does not replace the spatial and temporal domain of the current AHPS capabilities at the 13 River Forecast Centers (RFC) in areas known as `hydro-blind'. 
Furthermore, the NWM as of V1.2 has implemented not only assimilation of real-time United States Geological Survey (USGS) stage discharges but also assimilation of flow forecasts from the AHPS forecast points which are in turn routed downstream and updated whenever a new point is reached. 
This configuration of the NWM is known as `Replace and Route' or `RnR' and is used to enhance the forecasting skill of the NWM with the best available regional-scale data.
Figure \ref{ssec:national_water_model} shows Full-Resolution (FR) modeling stream network as well as the RnR mainstems (MS) network.

@@

As a result of high computational costs and large spatial domains, the need for high-resolution FIM at 10m or better requires additional post-processing from the principal output of the NWM which is forecast river discharges at the reach scale. The Height Above Nearest Drainage (HAND) terrain model is one such technique that can be used, along with synthetic rating curves (SRC), to convert riverine discharges to stages then finally to inundation extents.

@@

A core assumption made for HAND based FIM is enforcing drainage across the entire area of interest which requires significant DEM manipulations to make a reality.

@@

Due to significant advances in high-performance computing (HPC) and large scale high-resolution DEMs such as the National Elevation Dataset (NED) at the 10m scale, HAND has been implemented into software for large-scale, continental computation. 
HAND was initially implemented into operational software by the National Flood Interoperability Experiment (NFIE) to generate FIM hydrofabric (will be used interchangeably with the datasets produced by HAND) rapidly on a high-performance computer (HPC) \cite{maidment2017conceptual,liu2016cybergis}. 

@@

Originally, NFIE utilized the National Hydrography Dataset (NHD) Plus Medium Resolution (MR) to etch or burn flowlines prior to further conditioning but more recent work has advanced this to the more current NHDPlus High Resolution (HR) \cite{liu2020height}. 
The original NFIE dataset was employed by the NWC to produce forecast FIM from the NWM for use within its network of RFCs for additional guidance in hydro-blind regions.

@@

Overall, the software package is estimated to run CONUS at the full-resolution in 0.55 CPU years in a desktop setting.

Related to these, the USGS has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses synthetic rating curves with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{OWP FIM `Cahaba'}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Many of those in the academic community assessing HAND's efficacy for producing FIM have noted opportunities for improvement. 
\citeA{godbout2019error} found how reach length and slope are important parameters for maximizing mapping skill with the moderate values performing best. 
The co-linearity of reach length and slope led \citeA{godbout2019error} to propose that reaches of extreme lengths performed worse because of the extreme slope values, a parameter directly represented in Manning's equation. 
Issues with the reach-average approaches have been documented in \citeA{tuozzolo2019impact} where large within reach variance of the roughness Manning's n coefficient have been observed.
Furthermore, \citeA{garousi2019terrain} noted improvements to mapping efficacy by conditioning monotonically decreasing thalweg elevations, adjusting the Manning's n roughness coefficient, and using higher resolution (3m) Digital Elevation Model's (DEM) derived from light detection and ranging (Lidar).
The use of higher resolution DEMs in that study was motivated by previous work with Lidar DEMs and least-cost thalweg derivations \cite{zheng2018geoflood}.
Further work by \citeA{johnson2019integrated} noted the general under-prediction of HAND and suggested tuning the Manning's n parameter to better align SRC's with observations. 
Additionally, the sensitivity to low topographic relief and channel slope have been observed \cite{johnson2019integrated,godbout2019error}. 
Most notably, HAND has been shown to demonstrate sensitivity to drainage network density known colloquially as the catchment boundary problem \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
This sensitivity is noted at higher order streams with large flows and low Froude numbers and is manifested by a constriction in the inundation extents at the junction location.
These findings suggest improvements to HAND are required that utilize advanced computational algorithms and software to compute a FIM hydrofabric required for producing continental-scale FIM.

With all of the latest developments in the realm of continental FIM (CFIM) noted in the previous paragraph, a fast, open-source framework for accelerating the research to operations pipeline is required.
Here we introduce OWP FIM `Cahaba' that utilizes a few of the latest techniques in HAND based FIM oriented for use with NWM in continental scale operational forecasting settings. 
This framework is open-source and modular enabling and accelerating the research to operational development pipeline.
Automated evaluation tools and processed test case data enable the rapid testing and evaluation of new techniques in consistent contexts.
In addition to developed tooling, we introduce research demonstrating how FIM performance skill with HAND can be improved by reducing Horton-Strahler stream orders \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} of the stream networks.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to stream threshold which serves as a proxy for stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network to a singular stream order discretized by level paths, can enhance FIM skill.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM with applications.
%

>>>>

One of these operational systems is the Advanced Hydrologic Prediction System (AHPS) maintained by the National Oceanic Atmospheric Administration (NOAA) National Weather Service (NWS) with thousands of forecasting points across the US at typically short forecast horizons of 24 or 72 hours \cite{mcenery2005noaa}.
AHPS provides forecasting services in the form of ensemble stream flows at more than 3,600 locations and flood inundation maps (FIM) at more than 150 of those points shown in Figure \ref{fig:forecast_points}.

@@

Despite the AHPS advances in operational flood forecasting, it lacks sufficient spatial coverage and long-range forecast horizons to address the increasingly complex water challenges facing the US.

@@

\caption{Forecast points with and without FIM in United States' Advanced Hydrologic Prediction System.
Also show are the National Water Model stream networks at the full resolution (FR) and Mainstems (MS) resolution.}

@@

In response to growing stakeholder demand for enhanced and integrated water resource forecasts, the Office of Water Prediction (OWP) at the National Water Center (NWC) along with its partners has developed and implemented operationally the National Water Model (NWM) which is a version of the Weather Research and Forecast Hydrologic Model (WRF-Hydro) \cite{gochis2018wrf,cosgrove2019evolution}. 

@@

The NWM enhances the spatial and temporal domain of the current AHPS capabilities operated at the 13 River Forecast Centers (RFC) in areas known as `hydro-blind'.
As a complement to the operational NWM, OWP has also developed a configuration of the NWM that extends RFC forecasts downstream by assimilating and routing forecast streamflows to the next downstream AHPS forecast point.
This configuration of the NWM is used to enhance forecasting skill by leveraging best available regional-scale data.
The river network upon which this special configuration operates is herein referred to as the Main-Stem (MS) modeling stream network.
Figure \ref{ssec:national_water_model} shows the Full-Resolution (FR) modeling stream network as well as the MS network.

@@

As a result of high computational costs and large spatial domains, the need for high-resolution FIM at 10 m or better requires additional post-processing from the principal output of the NWM which is forecast river discharges at the reach scale. The Height Above Nearest Drainage (HAND) terrain model is one such technique that can be used, along with synthetic rating curves (SRC), to convert riverine discharges to stages then finally to inundation extents.

@@

A core assumption made for HAND based FIM is enforcing drainage across the entire area of interest which requires significant digital elevation maps (DEM) manipulations to make a reality.

@@

Due to significant advances in high-performance computing (HPC) and large scale high-resolution DEMs such as the 3D Elevation Program (3DEP) seamless at the 1/3 arc-second (10 m) scale, HAND has been implemented into software for large-scale, continental computation. 
As part of the OWP's Innovators Program and NWC's Summer Institute, the National Flood Interoperability Experiment (NFIE) generated FIM hydrofabric (will be used interchangeably with the datasets produced by HAND) rapidly on a high-performance computer (HPC) \cite{maidment2017conceptual,liu2016cybergis}. 

@@

Originally, NFIE utilized the NHD Medium Resolution (MR) to etch or burn flowlines prior to further conditioning but more recent work has advanced this to the more current NHDPlus High Resolution (HR) \cite{liu2020height}. 
The original NFIE dataset was employed by the NWC as an unofficial demonstrations to produce forecast FIM from the NWM for additional guidance in hydro-blind regions.

@@

Related to these efforts, the USGS has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses synthetic rating curves with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{OWP FIM `Cahaba'}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Many of those in the academic community assessing HAND's efficacy for producing FIM have noted opportunities for improvement. 
\citeA{godbout2019error} found how reach length and slope are important parameters for maximizing mapping skill with the moderate values performing best. 
The co-linearity of reach length and slope led \citeA{godbout2019error} to propose that reaches of extreme lengths performed worse because of the extreme slope values, a parameter directly represented in Manning's equation. 
Issues with the reach-average approaches have been documented in \citeA{tuozzolo2019impact} where large within reach variance of the roughness Manning's n coefficient have been observed.
Furthermore, \citeA{garousi2019terrain} noted improvements to mapping efficacy by conditioning monotonically decreasing thalweg elevations, adjusting the Manning's n roughness coefficient, and using higher resolution (3 m) Digital Elevation Model's (DEM) derived from light detection and ranging (Lidar).
The use of higher resolution DEMs in that study was motivated by previous work with Lidar DEMs and least-cost thalweg derivations \cite{zheng2018geoflood}.
Further work by \citeA{johnson2019integrated} noted the general under-prediction of HAND and suggested tuning the Manning's n parameter to better align SRC's with observations. 
Additionally, the sensitivity to low topographic relief and channel slope have been observed \cite{johnson2019integrated,godbout2019error}. 
Most notably, HAND has been shown to demonstrate sensitivity to drainage network density known colloquially as the catchment boundary problem \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
This sensitivity is noted at higher order streams with large flows and low Froude numbers and is manifested by a constriction in the inundation extents at the junction location.
These findings suggest improvements to HAND are required that utilize advanced computational algorithms and software to compute a FIM hydrofabric required for producing continental-scale FIM.

With all of the latest developments in the realm of continental FIM (CFIM) noted in the previous paragraph, a fast, open-source framework for accelerating the research to operations pipeline is required.
Here we introduce OWP FIM `Cahaba' that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
This framework is open-source and modular enabling and accelerating the research to operations development pipeline.
Automated evaluation tools and processed test case data enable the rapid testing and evaluation of new techniques in consistent contexts.
In addition to developed tooling, we introduce research demonstrating how FIM performance skill with HAND can be improved by reducing Horton-Strahler stream orders \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} of the stream networks.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to stream threshold which serves as a proxy for stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network to a singular stream order discretized by level paths, can enhance FIM skill.
This capability will be referred to as Generalized Mainstems (GMS).
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM with applications.
%
